Towards automated provenance collection for experimental runs of agent-based models\thanks{Supported by The Scottish Government}

### 2023-04-11 paper/paper.tex
Got the references correct. I need another 4 pages
content and some pretty diagrams.

Replication of social simulation results has been highlighted as a significant
issue for the ABM for a number of years (e.g.  \cite{edmonds2003replication}),
and in particular the paper that forms the basis of this work
\cite{polhill2017lessons}.  The focus of replication work has previous just
been on the model itself, but as was shown in ibid, the analysis of the
outputs of the model can potentially be just as complex, and no less difficult
to replicate unless adequate records are kept. Indeed there as an additional
attempt to reproduce the results and even with the lessons learnt from ibid, it
still took months to recreate the final diagrams that were submitted to paper.
The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
some guidance highlighting the need to keep a notebook of the analysis done and
a standardised approach to making that notebook; however, the lessons learned
from the replication exercise in \cite{polhill2017lessons} show that more
detailed guidance on the information that should be recorded.  and on tools
that could be used to support the process. Ideally the aim should be to
completely automate the process, record accurately sources of data used,
version check applications used essentially providing a complete graph from
data to result.

The output analysis replication in this paper concerns earlier work with
FEARLUS-SPOMM, which is a coupled agent-based model of agricultural decision-
making and species stochastic patch occupancy metacommunity model that has been
used to explore incentivisation strategies to improve biodiversity (Polhill et
al. 2013; Gimona and Polhill 2011). Belonging to the ‘typification’ class of
social simulations adfa, p. 1, 2011.  © Springer-Verlag Berlin Heidelberg 2011
(“theoretical constructs intended to investigate some properties that apply to
a wide range of empirical phenomena that share some common features” – Boero
and Squaz- zoni, 2005). This work involved the analysis of the outputs from
around 20,000 runs of the model using a number of techniques aimed at
demonstrating nonlinearities in the relationship between incentivisation and
biodiversity outcome.  Recording workflow data on the process used to create
analysis can be challenging, and currently there are no codified standards as
to how this should be done for ABMs. For FEARLUS-SPOMM, the methods used drew
heavily on statistical techniques available as R packages that are as part of
core R functionality. Although R allows transcripts of interactive terminal
sessions to be saved, the work involved great deal of exploration of different
ways of attempting to visualise and analyse the nonlineari- ties in the model
results, not all of which were likely to be reported in the paper. Such logs
are therefore not the best way to record the means by which the outputs were
analysed, and hence the strategy used was to save each analysis or
visualisation in a(n R) script. Since the output from the (Swarm) software that
generated the output data being analysed used a mixture of text formats, some
Perl scripts were also written to process that output into a CSV file for easy
import into R. When the MIRACLE pro- ject (Parker et al. 2015) provided a
context in which the replication of that analysis was necessary, an opportunity
was created to test the viability of the above strategy.

In the rest of this paper, we describe  a tool for automatically recording
metadata, which can be incorporated into the analysis replication process, and
how this  was used to regenerate some of the figures in Polhill et al. (2013)
and we test a tool to support keeping the records needed to per- form
replication more easily, before making some recommendations in concluding
remarks.

@@

ne of the aims was to leave as much of the code as unchanged as possible (this differs in approach to FAIR-CLI. develop this, both here and in the discussion.

One of the artefacts of the MIRACLE project \cite{} was the Social Simulation REplication Interface or SSREPI. This is the schema shown 
in \ref{fig:schema}. This schema has evolved since it original conception.
The newest version of this document may be found here \cite{}. This is a
database schema derived from the Dublin Core \cite{weibel2000dublin}, the
standard XML datatypes \cite{biron2004xml} and the PROV-O ontologies \cite{missier2013w3c}. The schema is designed with
agnosticisim towards the underlying database technology and as such has
been implmented both in PostgresSQL \cite{stonebraker1991postgres} and Sqlite3 \cite{sqliteorg2023syntax}.

@@

For the purposes of this short paper we are concentrating on a
demonstration of recording provenance. Indeed we use the example,
mentioned in the introduction of \cite{}, and modified the original Bash \cite{} scripts to include the what we denote as \textit{primitives} . So an example of the code

In order to record provenance we 
We coded the following primitives from
the SSREPI interface definition \cite{}. These are implemented in a mixture of Python \cite{} and Bash \cite{}

\begin{table}
\caption{Table captions should be placed above the
tables.}\label{tab1}
    \begin{tabularx}{\textwidth}{|l|p{2cm}|p{6cm}|}
\hline
Primitive & Type & Purpose \\
\hline
        SSREPI\_require\_minimum & Metadata & Lower bound on software hardware required \\ SSREPI\_require\_exact & Metadata & Exact bound on software hardware required \\
        SSREPI\_application & Provenance \& Metadata & specifies some executable \\
        SSREPI\_me & Provenance \& Metadata & Determines executable being run or returns a proper reference to the executable being run. \\
        SSREPI\_run & Provenance \& metadata & Blocking invocation of an executable which will allow the specification of all inputs, outputs and arguments. Creates run-time provenance information as well \\
        SSREPI\_batch & Provenance \& metadata & Non blocking invocation of an executable which will allow the specification of all inputs, outputs and arguments. Creates run-time provenance information as well \\
        SSREPI\_argument & Provenance & An argument type to an exectuable \\
        SSREPI\_output & Provenance & An output type from an executable \\
        SSREPI\_input & Provenance & An input type for an exectuable \\
        SSREPI\_hutton\_person & Metadata & Uses our institutions databases to populate metadata for a given individual \\
        SSREPI\_person & Metdata & Provide metadata for a particular actor within this system\\
        SSREPI\_project & Metadata & Specifies a project which contains all studies \\
        SSREPI\_study & Metadata & A set of experiments makes up a single study \\
        SSREPI\_set & Metdata & Sets the default licence and other metadata \\
        SSREPI\_involvement & Metadata & Links personnel to a study \\
        SSREPI\_paper & Metadata & A paper associated with this study \\
        SSREPI\_make\_tag & Metadata & Used for building a folksonomy \\
        SSREPI\_tag & Metadata & Used to tag any entity with a folksonomy tag \\
        SSREPI\_contributor & Metadata & A  person with some kind of relation to an executable or script. \\

@@

@@

@@

@@

What we have learnt from this 
This is a lot of pfaff. and is still not oven ready

So what use is this provenence meta data. Tracing bad data.

@@

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography[keyword={github.com/DougSalt/MABS2023}]{global.bib}
%
\end{document}

>>>>

Replication of social simulation results has been highlighted as a significant
issue for the ABM for a number of years (e.g.  \cite{edmonds2003replication}),
and in particular the paper that forms the basis of this work
\cite{polhill2017lessons}.  The focus of replication work has previously just
been on the model itself, but as was shown in ibid, the analysis of the
outputs of the model can potentially be just as complex, and no less difficult
to replicate unless adequate records are kept. Indeed there as an additional
attempt to reproduce the results and even with the lessons learnt from ibid, it
still took months to recreate the final diagrams that were submitted to paper.
The TRACE protocol \cite{schmolke2010ecological,ayllon2021keeping} provides
some guidance highlighting the need to keep a notebook of the analysis done and
a standardised approach to making that notebook; however, the lessons learned
from the replication exercise in \cite{polhill2017lessons} show that more
detailed guidance on the information that should be recorded,  and on tools
that could be used to support the process. Ideally the aim should be to
completely automate the process, record accurately sources of data used,
version check applications used essentially providing a complete graph from
data to result.

The output analysis replication in this paper concerns earlier work with
FEARLUS-SPOMM, which is a coupled agent-based model of agricultural decision-
making and species stochastic patch occupancy metacommunity model that has been
used to explore incentivisation strategies to improve biodiversity
\cite{polhill_nonlinearities_2013}, \cite{gimona2011exploring}. Belonging to the
‘typification’ class of social simulations  \cite{boero2005does}.  This work
involved the analysis of the outputs of approximately 20,000 runs of the model
using a number of techniques aimed at demonstrating nonlinearities in the
relationship between incentivisation and biodiversity outcome.  Recording
workflow data on the process used to create analysis can be challenging, and
currently there are no codified standards as to how this should be done for
ABMs. For FEARLUS-SPOMM, the methods used drew heavily on statistical
techniques available as R packages that are as part of core R functionality.
Although R allows transcripts of interactive terminal sessions to be saved, the
work involved great deal of exploration of different ways of attempting to
visualise and analyse the nonlinearities in the model results, not all of
which were likely to be reported in the paper. Such logs are therefore not the
best way to record the means by which the outputs were analysed, and hence the
strategy used was to save each analysis or visualisation in a(n R) script.
Since the output from the (Swarm) software that generated the output data being
analysed used a mixture of text formats, some Perl scripts were also written to
process that output into a CSV file for easy import into R. When the MIRACLE pr
ject \cite{parker2019final} provided a context in which the replication of that
analysis was necessary, an opportunity was created to test the viability of the
above strategy.

In the rest of this paper, we describe  a tool for automatically recording
metadata, which can be incorporated into the analysis replication process, and
how this  was used to regenerate some of the figures in
\cite{polhill_nonlinearities_2013} and we test a tool to support keeping the
records needed to per- form replication more easily, before making some
recommendations in concluding remarks.

@@

One of the artefacts of the MIRACLE project \cite{parker2019final} was the
Social Simulation REplication Interface or SSREPI. This is the schema shown in
\ref{fig:schema}.  This schema has evolved since it original conception.  The
newest version of this document may be found here \cite{polhill2022miracle}. This is a database
schema derived from the Dublin Core \cite{weibel2000dublin}, the standard XML
datatypes \cite{biron2004xml} and the PROV-O ontologies \cite{missier2013w3c}.
The schema is designed with agnosticisim towards the underlying database
technology and as such has been implmented both in PostgresSQL
\cite{stonebraker1991postgres} and Sqlite3 \cite{sqliteorg2023syntax}.

One of the aims was to leave as much of the code, i.e. the original experiment
as unchanged as possible. That is we wanted to change the framework code rather
than the core code. The core code in the example above being the executable,
precompiled in Objective-C \cite{objectivec}. Additionally the support code, i.e. the
code that prepared the data, a Perl \cite{perl563} script and the code that did the
post analysis, a mixture of Perl and R \cite{R422}.  This meant we developed the
database access in Python \cite{python368}, wrote wrapper scripts in Bash \cite{bash4420}. This
differs in approach to FAIR data piperline \cite{mitchell2022fair} in which the provenance framework is
embedded in the code that constitutes the body of the experiment as it were.

With the tool in the current state this involved labouriously going through the
code and inserting the some of the following Bash-based commands, shown in
table \ref{tab:commands}. These commands have a fairly complex parameter set as
it stands, and no way of checking whether the code worked other than running
it, and this led to one of the immediate problems with this method.

@@

For the purposes of this short paper we are concentrating on a
demonstration of recording provenance. Indeed we use the example,
mentioned in the introduction of \cite{polhill2022miracle}, and modified the original Bash \cite{bash4420} scripts to include the what we denote as \textit{primitives} . So an example of the code

In order to record provenance we 
We coded the following primitives from
the SSREPI interface definition \cite{polhill2022miracle}. These are implemented in a mixture of Python and Bash.

\begin{table}
\caption{Table captions should be placed above the
tables.}\label{tab:commands}
    \begin{tabularx}{\textwidth}{|l|p{2cm}|p{6cm}|}
\hline
Primitive & Type & Purpose \\
\hline
        SSREPI\_require\_minimum & Metadata & Lower bound on software hardware required \\ SSREPI\_require\_exact & Metadata & Exact bound on software hardware required \\
        SSREPI\_application & Provenance \& Metadata & specifies some executable \\
        SSREPI\_me & Provenance \& Metadata & Determines executable being run or returns a proper reference to the executable being run. \\
        SSREPI\_run & Provenance \& metadata & Blocking invocation of an executable which will allow the specification of all inputs, outputs and arguments. Creates run-time provenance information as well \\
        SSREPI\_batch & Provenance \& metadata & Non blocking invocation of an executable which will allow the specification of all inputs, outputs and arguments. Creates run-time provenance information as well \\
        SSREPI\_argument & Provenance & An argument type to an exectuable \\
        SSREPI\_output & Provenance & An output type from an executable \\
        SSREPI\_input & Provenance & An input type for an exectuable \\
        SSREPI\_hutton\_person & Metadata & Uses our institutions databases to populate metadata for a given individual \\
        SSREPI\_person & Metdata & Provide metadata for a particular actor within this system\\
        SSREPI\_project & Metadata & Specifies a project which contains all studies \\
        SSREPI\_study & Metadata & A set of experiments makes up a single study \\
        SSREPI\_set & Metdata & Sets the default licence and other metadata \\
        SSREPI\_involvement & Metadata & Links personnel to a study \\
        SSREPI\_paper & Metadata & A paper associated with this study \\
        SSREPI\_make\_tag & Metadata & Used for building a folksonomy \\
        SSREPI\_tag & Metadata & Used to tag any entity with a folksonomy tag \\
        SSREPI\_contributor & Metadata & A  person with some kind of relation to an executable or script. \\

@@

As mentioned the method undertaken in this paper involved inserting the correct
Bash-based commands, listed in table \ref{tab:commands}. These commands
have a fairly complex parameter set as it stands, and no way of checking
whether the code worked other than running it, and this led to one of the
immediate problems with this method. The only way to see if the code was
working was to run it, a suck it and see methodology. This was extremely time consuming. We took the obvious approach and ran small subsets of
the ~20,000 runs or so, to see if we could get the framework to work and then scaled it up to the full run, but invariably we ran into new problems the greater the scale. It therefore took a substantial time to even get this first replication to run. It took several months of coding effort to get this correct.

@@

So what use is this provenence meta data? Beyond replicability

@@

Those with a sharp eye may have noticed the versions of Perl, Python and R were not around when the original experiments were run and obviously this has implications.

@@

What we have learnt from this?  As noted the process of running the experiment with the new provenance and metadata commands was a complete pfaff. It involved the repeated running of the experiment with a larger and larger array of inputs until we matched the size of the original experiment.  There were codes in the both the Python and bash code (calling the Python scripts)

This has taken me almost a year to get it running. Repeating an experiment in a completely automated fashion turns out tbe a really hard task. The plan is to adapt this provenance framework to other model running language frameworks, in particular R and Python.

@@

The underlying database is probably wrong. We should be using a graphing language such as Gremlin

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
\bibliographystyle{splncs04}
\bibliography{global.bib}
% \bibliography[keyword={github.com/DougSalt/SSC2023}]{global.bib}
%
\end{document}
